\documentclass[12pt,a4paper]{cibb}

% Fallback definition for missing \@ordinalM
\makeatletter
\providecommand{\@ordinalM}[2]{#1}
\makeatother

\usepackage{subfigure,graphicx}
\usepackage{amsmath,amsfonts,latexsym,amssymb,euscript,xr}
\usepackage{booktabs}
\usepackage[nodayofweek]{datetime}
\usepackage{hyperref}
\usepackage{fmtcount}
\usepackage[english]{datenumber}
\usepackage[absolute]{textpos}

\usepackage[table]{xcolor}
\usepackage{color,colortbl,tabularx}

\usepackage[english]{babel}
\usepackage[protrusion=true,expansion=true]{microtype}
\usepackage{amsmath,amsfonts,amsthm}
\usepackage{pifont}

\def\red{\color{red}}
\def\black{\color{black}}
\def\blue{\color{blue}}
\def\magenta{\color{magenta}}

\definecolor{LightBlue}{rgb}{0.88,0.9,0.9}

\newcommand\blfootnote[1]{%
  \begingroup
  \renewcommand\thefootnote{}\footnote{#1}%
  \addtocounter{footnote}{-1}%
  \endgroup
}

\title{\Large $\ $\\ \bf Phone-Watch Contrastive Representation Learning for Subject-Generalized Wearable Activity Recognition}

\author{\blue \large First Author$^1$, Second Author$^{2}$, and Third Author$^{*,3,4}$}
\address{\blue \footnotesize $\ $\\$^1$ First author's department, institute,
city, country. \\
%
$^2$ Second author's department, institute,
city, country. \\
%
$^3$ Third author's primary affiliation department, institute,
city, country.  \\
%
$^4$ Third author's secondary affiliation department, institute,
city, country. \\
%
% \bigskip
% ORCID codes: FA 0000-0000-0000-0000; SA 0000-0000-0000-0000; TA 0000-0000-0000-0000.
% \bigskip
% \newline
% $^*$corresponding author: email@email.com
% \bigskip
% \newline
% [The initial submission must be anonymous.
% Do not edit the blue text sections of this short paper template before the initial submission.
% Include information about author names and affiliations, conflict of interests, acknowledgments, data and software availability, and funding in black text only in the resubmission.]
}

\abstract{\small wearable sensing, human activity recognition, contrastive learning, multimodal fusion, subject generalization. \normalsize
\\[17pt]
{\bf Abstract (max 250 words).} This report studies activity recognition from synchronized phone and watch inertial streams using a subject-disjoint evaluation setup.
The raw dataset provides per-subject files for phone and watch accelerometer and gyroscope recordings, with activity labels stored as contiguous letter-coded segments.
We first build a segment-aware preprocessing pipeline that pairs the four streams by activity episode, rescales them onto a shared 20 Hz timeline, and writes normalized 3 s windows with 1 s stride.
This produces 50,091 training windows, 12,744 validation windows, and 47,790 test windows, each represented as a $(12, 60)$ tensor.

We then compare two learning strategies.
The first is a supervised 1D CNN fusion baseline trained directly on the 12-channel windows.
The second is a phone-watch contrastive model that learns paired embeddings with a symmetric InfoNCE loss and is evaluated through frozen linear probes.
The canonical paired probe improves over the fused supervised baseline at both 100\% and 10\% label budgets, reaching 48.61\% vs.\ 44.81\% test accuracy with full labels and 47.95\% vs.\ 42.43\% with 10\% labels.
The gain is larger under limited labels, which supports the claim that cross-view pretraining improves label efficiency.

An important side result is strong device asymmetry.
Watch-only models are much stronger than phone-only models, which suggests that wrist motion carries most of the discriminative signal in the current setup.
This draft records the pipeline, the current quantitative findings, and the main open questions to address with later figures and visualization.
}

\begin{document}

\renewcommand{\thefootnote}{}
\footnotetext{\small{Article version: \datedate $\;$ h\currenttime  $\;$ CET}}

\thispagestyle{myheadings}
\pagestyle{myheadings}
\markright{\tt Proceedings of CS 2026}

\section{Introduction}
\label{sec:SCIENTIFIC-BACKGROUND}

Human activity recognition from wearables is a natural test bed for representation learning because labels are expensive while unlabeled sensor streams are plentiful.
Our project focuses on a phone-watch setting where the two devices observe the same activity from different body locations.

The working hypothesis is that synchronized phone and watch windows should provide complementary views of the same underlying motion pattern.
If we align those views during pretraining, the learned representation should transfer better to unseen subjects and low-label regimes.

This report therefore has two goals.
The first is to document a reliable preprocessing and windowing pipeline for the four sensor streams.
The second is to compare a supervised fusion baseline against cross-view contrastive pretraining followed by frozen linear probes.

\section{Dataset Exploration}
\label{sec:DATA}

\subsection{Raw Dataset Structure}

The raw data are organized into subject-disjoint \texttt{train/} and \texttt{test/} folders.
Each split contains four sensor collections: \texttt{phone/accel}, \texttt{phone/gyro}, \texttt{watch/accel}, and \texttt{watch/gyro}.

Every file stores one subject and one stream.
Each row contains \texttt{subject\_id}, an activity label, a timestamp, and three sensor axes.
The repository currently exposes labels as letter codes rather than human-readable activity names.

The cached window manifest contains 18 activity codes: A, B, C, D, E, F, G, H, I, J, K, L, M, O, P, Q, R, and S.
Because the semantic mapping is not yet stored in the repo, all per-class analysis in this draft uses those codes directly.

\subsection{Split Summary}

The raw paired-segment index contains 628 candidate activity episodes.
Of those, 625 episodes have all four required streams and 3 are dropped because the phone accelerometer and phone gyroscope are missing.

After preprocessing, we obtain the subject-disjoint split shown in Table~\ref{tab:split-summary}.
Validation subjects are deterministically carved out of the raw training subjects, so the final workflow uses train, validation, and test subjects without leakage.

\begin{table}[httb!] \small
\centering
    \begin{tabularx}{0.9\textwidth}{ l  c  c  X }
        \toprule
        \textbf{Split} & \textbf{\# Subjects} & \textbf{\# Windows} & \textbf{Notes} \\
        \midrule
        \rowcolor{LightBlue} Train & 16 & 50,091 & Used for supervised fitting and contrastive pretraining. \\
        Val & 4 & 12,744 & Held out from the original training subjects for model selection. \\
        \rowcolor{LightBlue} Test & 14 & 47,790 & Unseen subjects only; subject 1629 contributes a second activity occurrence for every class. \\
        \bottomrule
    \end{tabularx}
    \caption{\textbf{Window cache summary}.
    The cached windows under \texttt{artifacts/windows/} are the source of truth for all current experiments.
    \label{tab:split-summary}}
\end{table}

The split is close to class-balanced.
Validation and test have 708 and 2,655 windows per class, respectively.
Training is nearly balanced, with small deficits for B, F, J, and O because three raw activity episodes were removed before caching.

\section{Annotation and Signal Exploration}

The dataset is not best interpreted as one continuous, globally synchronized recording per subject.
Instead, each file is an ordered concatenation of contiguous activity episodes, and the same label can appear multiple times for one subject.

For that reason, we index episodes by \texttt{(subject\_id, activity\_label, occurrence\_index)} rather than by absolute time alone.
This segment-aware view is the key design choice that keeps phone and watch alignment faithful to the annotation structure.

The valid paired episodes are highly regular.
Their common overlapping duration is approximately 179.71 s on average, with only a very small spread across subjects and classes.
With a 3 s window and 1 s stride, each standard episode yields 177 windows.

The native sampling rates differ noticeably across streams.
Across valid segments, phone accelerometer averages 25.55 Hz, phone gyroscope 21.14 Hz, and both watch streams about 20.79 Hz, with some files much faster.
This heterogeneity makes direct sample-by-sample fusion inappropriate without resampling.

\section{Signal Preprocessing}

The preprocessing pipeline begins by parsing each stream into contiguous runs of the same activity code.
Each run becomes one candidate segment with start row, end row, duration, and native sampling-rate metadata.

We then build four-stream episode pairs by matching segments with the same activity code and occurrence index.
Only episodes that contain phone accelerometer, phone gyroscope, watch accelerometer, and watch gyroscope are kept.

Within every valid episode, we compute the common overlapping duration across the four streams.
All later operations use relative time within that overlap rather than raw timestamps from the original files.

Each sensor axis is linearly interpolated onto a shared 20 Hz grid.
The final fused tensor stacks the streams in a fixed order and produces a normalized $(12, 60)$ window.
Channel-wise z-score statistics are estimated on the training windows and stored in \texttt{artifacts/windows/manifest.json}.

We intentionally do not add handcrafted denoising, spectral transforms, or manual feature engineering at this stage.
The current comparison isolates the value of alignment, normalization, and representation learning.

\section{Windowing Strategies}

Window generation is centered on a simple fixed-rate policy.
Each window spans 3.0 s, uses a 1.0 s stride, and contains 60 samples after resampling.

Window centers range from 1.5 s to 177.5 s inside a standard episode.
Because windows are generated only after segment pairing, each window belongs to a single activity episode and does not cross label transitions.

This segment-first design removes the need for majority-vote labeling or ad hoc transition filtering.
In effect, the labels are inherited from the paired episode, and every cached window is already aligned across phone and watch.

For label-efficiency experiments, the 10\% subset is selected deterministically within each \texttt{(subject, class)} bucket.
That policy keeps at least one example per bucket when possible and ensures that the 10\% supervised baseline and the 10\% probe use comparable labeled subsets.

\section{Feature Extraction}

The current system does not rely on handcrafted statistical features.
Instead, it learns features directly from the normalized time-domain windows.

The supervised baseline uses a 1D CNN encoder with three temporal convolution blocks, batch normalization, ReLU activations, max pooling, and global average pooling.
The encoder outputs a 256-dimensional feature vector and a linear classifier maps that feature to one of the 18 activity codes.

The contrastive model uses separate phone and watch encoders with the same backbone.
Each encoder consumes a $(6, 60)$ view and produces a 256-dimensional representation.
Projection heads map those features from 256 to 128 to 64 dimensions before the contrastive loss is applied.

For downstream evaluation, the paired probe concatenates the frozen phone and watch embeddings into a 512-dimensional feature vector.
Single-device probes reuse only the phone or watch embedding, which helps isolate how much each device contributes.

\section{Modeling and Results}

\subsection{Supervised and Contrastive Setup}

The supervised baseline is trained end to end with cross-entropy loss.
The default optimizer is AdamW with learning rate $10^{-3}$, weight decay $10^{-4}$, batch size 256, dropout 0.2, and early stopping on validation macro F1.

Contrastive pretraining uses the training windows without labels.
Phone and watch views are augmented with Gaussian jitter, mild amplitude scaling, and short temporal masking before they are passed through the dual encoders.

The pretraining objective is a symmetric InfoNCE loss with in-batch negatives and temperature $\tau = 0.2$:
\begin{equation}
\mathcal{L} =
\frac{1}{2N}
\sum_{i=1}^{N}
\left[
- \log \frac{\exp(\mathrm{sim}(z_i^p, z_i^w)/\tau)}{\sum_{j=1}^{N}\exp(\mathrm{sim}(z_i^p, z_j^w)/\tau)}
- \log \frac{\exp(\mathrm{sim}(z_i^w, z_i^p)/\tau)}{\sum_{j=1}^{N}\exp(\mathrm{sim}(z_i^w, z_j^p)/\tau)}
\right].
\label{eq:infonce}
\end{equation}

After pretraining, the encoder weights are frozen.
We then train linear heads for paired, phone-only, and watch-only probes at 10\% and 100\% label fractions.

\subsection{Evaluation Protocol}

All headline numbers are reported on the held-out test subjects.
We track accuracy, macro F1, per-class F1, per-subject accuracy, confusion matrices, and per-subject CSV summaries.

The most important comparison is the fused supervised baseline versus the paired frozen probe.
That comparison isolates whether cross-view pretraining gives a useful representation before any end-to-end fine-tuning.

\section{Results}
\label{sec:RESULTS}

\subsection{Quantitative Summary}

Table~\ref{tab:main-results} summarizes the current results.
The paired contrastive probe outperforms the canonical fusion baseline at both label budgets.
At full labels, accuracy improves from 0.4481 to 0.4861 and macro F1 from 0.4395 to 0.4707.

The low-label comparison is even stronger.
With only 10\% of the labeled training windows, the fused supervised baseline reaches 0.4243 accuracy, while the paired probe reaches 0.4795.
That is a gain of 5.52 accuracy points and 6.13 macro-F1 points.

\begin{table}[httb!] \small
\centering
    \begin{tabularx}{0.96\textwidth}{ l  l  c  c  c }
        \toprule
        \textbf{Method} & \textbf{Input} & \textbf{Labels} & \textbf{Acc.} & \textbf{Macro F1} \\
        \midrule
        \rowcolor{LightBlue} Supervised fusion baseline & phone + watch raw windows & 100\% & 0.4481 & 0.4395 \\
        Supervised fusion baseline & phone + watch raw windows & 10\% & 0.4243 & 0.4045 \\
        \rowcolor{LightBlue} Contrastive paired probe & frozen phone + watch embeddings & 100\% & 0.4861 & 0.4707 \\
        Contrastive paired probe & frozen phone + watch embeddings & 10\% & 0.4795 & 0.4658 \\
        \rowcolor{LightBlue} Supervised watch baseline & watch raw windows only & 100\% & 0.6355 & 0.6391 \\
        Contrastive watch probe & frozen watch embeddings only & 100\% & 0.5831 & 0.5777 \\
        \rowcolor{LightBlue} Supervised phone baseline & phone raw windows only & 100\% & 0.1758 & 0.1712 \\
        Contrastive phone probe & frozen phone embeddings only & 100\% & 0.1692 & 0.1540 \\
        \bottomrule
    \end{tabularx}
    \caption{\textbf{Current headline results}.
    The paired probe beats the canonical fusion baseline under both full-label and 10\%-label training.
    \label{tab:main-results}}
\end{table}

\subsection{Interpretation}

The strongest single observation is device asymmetry.
The watch-only supervised baseline reaches 0.6355 accuracy, far above both the fusion baseline and any phone-only model.
This suggests that wrist motion is currently the cleanest source of discriminative information.

The second observation is that paired contrastive pretraining is useful even though the downstream head is linear and the encoders are frozen.
The minimum per-subject test accuracy rises from 0.1946 in the full fusion baseline to 0.2307 in the full paired probe, and from 0.1111 to 0.2341 in the 10\% comparison.

Per-class behavior is also consistent across runs.
Classes B, O, and P are among the easiest, while F, I, J, K, and L remain difficult.
Because the class semantics are not yet included in the repository, we defer behavior-level interpretation until the label mapping is added.

An additional exploratory run with a low-learning-rate paired setup reaches about 0.501 test accuracy and 0.486 macro F1.
We treat that result as promising headroom rather than the main claim, because the standardized comparison table is still based on the canonical artifact set.

\subsection{What Still Needs Figures}

The repo already saves confusion matrices, per-subject accuracy CSV files, and walkthrough plots under \texttt{artifacts/}.
Those should become the final figures for this report once we decide on the exact visual style.

The highest-priority additions are a split-summary figure, one example of segment alignment and resampling, confusion matrices for the fusion baseline and paired probe, and a per-subject accuracy comparison plot.
Embedding visualizations can be added later once we expand the representation-analysis story.

\section{Conclusion}
\label{sec:CONCLUSIONS}

This project now has a complete end-to-end pipeline from raw phone and watch inertial files to subject-disjoint quantitative evaluation.
The main engineering contribution is a segment-aware alignment and resampling pipeline that converts heterogeneous raw streams into stable multimodal windows.

The main empirical result is that cross-view contrastive pretraining improves over the canonical fused supervised baseline, especially when labels are limited.
At the same time, the experiments reveal a strong watch-over-phone advantage that should shape the next round of analysis.

The next steps are clear.
We should add the human-readable label mapping, integrate the saved plots into the paper, and study why watch-only models outperform fusion in the current implementation.
We should also test whether stronger contrastive checkpoints and light fine-tuning can close more of the gap to the watch-only baseline.

\section*{Availability of data and software code (optional and strongly suggested)}
\label{sec:AVAILABILITY}

The raw inertial files used in this draft are stored in \texttt{data/}.
Derived window caches are stored in \texttt{artifacts/windows/}, supervised outputs in \texttt{artifacts/baseline/}, contrastive outputs in \texttt{artifacts/contrastive/}, and probe outputs in \texttt{artifacts/probes/}.

Reusable training and evaluation code is implemented under \texttt{src/}, and the current execution workflow is documented in \texttt{docs/phase2\_runbook.md}.
Before public release, we should confirm whether the raw dataset can be redistributed directly or only referenced.

\footnotesize
\bibliographystyle{unsrt}
\bibliography{bibliography}
\normalsize

\end{document}
